A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-25 _manuscript/_tex/manuscript.tex
update _freeze and _manuscript

\authorsaffiliations{
{Department of Psychology, LMU Munich},{Institute of Behavioral Science
\& Technology, University of St.~Gallen},{Department of Business
Psychology, Technical University of Applied Sciences Augsburg},{Global
Business School for Health, University College London}}

@@

\abstract{When planning experimental research, determining an
appropriate sample size and using suitable statistical models are
crucial for robust and informative results. The recent replication
crisis underlines the need for more rigorous statistical methodology and
adequately powered designs. Generalized linear mixed models (GLMMs)
offer a flexible statistical framework to analyze experimental data with
complex (e.g., dependent and hierarchical) data structures. However,
available methods and software for a priori sample size planning for
GLMMs are often limited to specific designs. Tailored data simulation
approaches are a more flexible alternative. Based on a practical case
study, the current tutorial equips researchers with a step-by-step guide
and corresponding code for conducting tailored a priori sample size
planning with GLMMs. We not only focus on power analysis but also
explain how to use the precision of parameter estimates to determine
appropriate sample sizes. We conclude with an outlook on the increasing
importance of simulation-based sample size planning.}

@@

\authornote{\par{\addORCIDlink{Florian
Pargent}{0000-0002-2388-553X}}\par{\addORCIDlink{Timo K.
Koch}{0000-0001-6728-2063}}

\par{ This is version 2.5 of our preprint published at
\url{https://doi.org/10.31234/osf.io/rpjem}. All materials (reproducible
manuscript, including R code and all simulation results) are available
in the project's repository on the Open Science Framework (OSF) at
\url{https://osf.io/dhwf4/}. A Quarto Manuscripts website is hosted at
\url{https://Timo-Ko.github.io/glmm_simulation_tutorial/}.  The authors
declare that there were no conflicts of interest with respect to the
authorship or the publication of this article. This research was funded
by a grant from the Volkswagen Foundation (Grant
No.~98525).   Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: Florian
Pargent:   formal
analysis, methodology, visualization, writing, editing; Timo K.
Koch:   conceptualization, formal
analysis, writing, editing; Anne-Kathrin Kleine:   formal
analysis, editing; Eva Lermer:   conceptualization, funding
aquisition, editing; Susanne
Gaube:   conceptualization, supervision, editing}
\par{Correspondence concerning this article should be addressed to Timo
K. Koch, Institute of Behavioral Science \& Technology, University of
St.~Gallen, Torstrasse
25, St.~Gallen 9000, Switzerland, Email: timo.koch@unisg.ch}
}

@@

When planning experimental research, it is essential to determine an
appropriate sample size and to use appropriate statistical models to
analyze the data to ensure that the results obtained are both robust and
informative (\citeproc{ref-lakensSampleSizeJustification2022}{Lakens,
2022a}). The recent replication crisis has illustrated many challenges
surrounding the reproducibility and reliability of study findings
(\citeproc{ref-yarkoniGeneralizabilityCrisis2022}{Yarkoni, 2022}). As a
result, there is a growing need for more rigorous statistical
methodology and the adoption of adequately powered experimental designs.
Multiple easy-to-use software solutions exist for simple statistical
models and experimental designs. However, many researchers lack the
skills and tools to conduct ``a priori'' (i.e., before data collection)
sample size planning for more complex research designs such as flexible
generalized linear mixed models (GLMM) framework. In the present work,
we provide a tutorial on how to determine adequate sample sizes by
performing tailored simulation-based sample size planning for GLMMs.
After introducing some theoretical background on sample size planning,
we review existing software solutions in R and discuss under which
circumstances tailored data simulations are necessary. Then we describe
general steps and decisions involved in tailored data simulation. To
illustrate the details of these steps, we finish with a hypothetical
case study from the field of human-AI (artificial intelligence)
interaction research.

To benefit most of this tutorial paper, we recommend readers to
familiarise themselves with basic statistical concepts like hypothesis
tests (HTs) and their statistical power as well as confidence intervals
(CIs) and their precision
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021};
\citeproc{ref-lakensImprovingYourStatistical2022}{Lakens, 2022b};
\citeproc{ref-riesthuisSimulationBasedPowerAnalyses2024}{Riesthuis,
2024}). Some knowledge of causal inference is useful but not necessary
(\citeproc{ref-deffnerCausalFrameworkCrossCultural2022}{Deffner et al.,
2022}; \citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al.,
2021}). In addition, readers should have a rough understanding of R
(\citeproc{ref-wickhamDataScienceImport2023}{Wickham et al., 2023}) and
how to simulate data. For data simulation, we use functions from the
tidyverse (\citeproc{ref-wickhamWelcomeTidyverse2019}{Wickham et al.,
2019}) and the faux package (\citeproc{ref-R-faux}{DeBruine, 2023}).
Finally, readers should be familiar with regression modeling in general
and GLMMs in particular. In this tutorial, we simulate data by manually
specifying the model equation of a GLMM that represents our assumed data
generating process
(\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}). It is not necessary to understand the technical details of
how GLMMs are estimated. However it is crucial to understand the
structure of a basic GLMM (e.g., logistic regression with random
intercepts) and how the model assumes that the values in the dependent
variable are determined by the predictor variables and the random
effects.

@@

Conducting research with insufficiently large sample sizes can have many
negative consequences. First, experiments may yield inconclusive or
misleading results, hindering the accumulation of knowledge. Second,
studies that are doomed to never finding a postulated effect waste
resources by consuming time, effort, and funding without delivering
meaningful results. For these reasons, many journals and funding bodies
now require that a justification of sample size is included in study
protocols and grant proposals, recognizing its significance in ensuring
robust and meaningful findings. Although scientists often do not justify
sample size or use general heuristics from the literature, resource
constraints often play an important role
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}). But
ideally, a suitable sample size should be determined a priori (i.e.,
before the study is conducted) based on some meaningful computation to
ensure that the study will be able to fulfill its purpose.

The majority of empirical studies in psychology and other social
sciences apply hypothesis testing. As a consequence, the dominant
approach for determining an adequate sample size is based on power
analysis (i.e., planning for power)
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}).
Statistical power is defined as the probability that a HT has a
significant p-value when analyzing repeated samples from a population
with a true effect of some pre-specified size
(\citeproc{ref-cohenPowerPrimer1992}{Cohen, 1992}). Less formally, power
is described as the probability that a HT correctly rejects the null
hypothesis when the alternative hypothesis is true. If the sample size
(i.e., the number of participants and/or stimuli) used for data
collection is insufficient to detect the effects or relationships being
investigated with high probability, the study is considered
``underpowered''. When planning for power, a target is set for the
statistical power of a HT of interest. Assuming some effect size of
interest and a desired significance level, a minimum sample size can be
determined that, on average, would guarantee reaching this target.
Although it is recommended to justify the desired significance level and
power (\citeproc{ref-lakensJustifyYourAlpha2018}{Lakens, Adolfi, et al.,
2018}), most empirical studies adopt the heuristic of \(\alpha = 0.05\)
and \(1 - \beta = 0.80\).

In contrast to power analysis, sample size planning can also be based on
the precision of parameter estimates (i.e., planning for precision or
planning for accuracy)
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}). Not
all research questions are best answered using hypothesis testing. It
has been argued that basic research rarely requires making discrete
decisions on whether some effect has been ``discovered'' and should thus
shift from a hypothesis testing towards an estimation framework
(\citeproc{ref-cummingNewStatisticsWhy2014}{Cumming, 2014};
\citeproc{ref-kruschkeBayesianNewStatistics2018}{Kruschke \& Liddell,
2018};
\citeproc{ref-mcelreathStatisticalRethinkingBayesian2020}{McElreath,
2020}). Although this view is not without critique
(\citeproc{ref-uyguntuncEpistemicPragmaticFunction2023}{Uygun Tunç et
al., 2023}), at least for exploratory or pilot studies where little
previous research has been conducted, more scientists seem to agree that
simply estimating the effects of interest and making the estimation
uncertainty transparent by reporting CIs is more useful. Assuming that
no HTs are conducted for a planned study, power analysis is not relevant
for sample size planning. Nonetheless, the sample size still has a
crucial effect on how informative the planned study will be, because an
effect of interest is estimated more precisely with bigger samples. In
the precision framework, the target quantity commonly used for sample
size planning is the expected width of a CI
(\citeproc{ref-kelleySampleSizePlanning2006}{Kelley \& Rausch, 2006};
\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}). The
values inside a CI are often interpreted as plausible values for the
quantity of interest it is supposed to estimate. More formally, a CI
with confidence level 0.95 provides the smallest interval with the
property that upon repeated sampling, 95\% of individual CIs would
include the true quantity of interest. Thus, a narrow CI is more
informative about the size of the true effect than a wide interval.
Apart from the desired confidence level, the width of a CI depends
strongly on the sample size. Because bigger samples carry more
information, they lead to smaller CIs. When planning for precision, a
target can be set for the expected width of a CI of interest. Assuming
some effect size of interest and a desired confidence level, a minimum
sample size can be determined that, on average, would guarantee reaching
this target. Because planning for precision is still rare, there are no
common heuristics on how to choose the desired width of the CI
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).

@@

As study designs become more complex, psychological researchers require
more sophisticated statistical models to capture the nuanced
relationships and grouping structures introduced by them
(\citeproc{ref-yarkoniGeneralizabilityCrisis2022}{Yarkoni, 2022}). GLMMs
(also called multilevel models) are gaining increasing popularity
because they offer great flexibility
(\citeproc{ref-fahrmeirRegressionModelsMethods2021}{Fahrmeir et al.,
2021}). GLMMs are an extension of LMMs (Linear Mixed Models), which are,
in turn, extensions of linear regression models that account for
correlated data, including hierarchical structures
(\citeproc{ref-fahrmeirRegressionModelsMethods2021}{Fahrmeir et al.,
2021}). In this context, correlated data means that the value in the
outcome variable for one observation may be related to the value of
another observation in a systematic way that is not already accounted
for by the usual (fixed) predictor variables (e.g., age of
participants). This correlation can arise for various reasons: For
instance, responses to some stimuli from some participants might be more
similar because the same person was measured twice (repeated
measurements), participants come from the same neighborhood
(clustering), or participants responded to the same stimulus (stimulus
effects). Thus, modeling such correlations is important whenever the
data has a clear structure, while the grouping variables can be
hierarchically nested (e.g., grouping variables students and schools:
each student belongs to exactly one school) or cross-classified (e.g.,
grouping variables students and math exercises: each student is
presented with several math exercises). LMMs are used when the outcome
variable is continuous and follows a normal distribution. They allow for
the modeling of fixed effects, which capture the relationships between
the usual predictors and the outcome, as well as random effects, which
account for the different types of correlation structure and grouping
effects. Random effects are typically assumed to follow a normal
distribution with a mean of zero and a variance that quantifies the
heterogeneity across groups. GLMMs extend the LMM framework to
accommodate non-normally distributed continuous and categorical outcome
variables. GLMMs involve a link function that connects the linear
combination of predictor variables to the expected value of the outcome
variable. The link function allows for modeling the relationship between
predictors and the outcome in a non-linear way that is appropriate for
the specific distribution family of the outcome variable. For example,
think of an experiment with different design factors (e.g., picture
positions, headline aesthetics) impacting the likelihood of users
clicking on an online advertisement. The participants' behavior is
measured repeatedly over several sessions. The click patterns of
participants in one session are likely to be correlated with their
previous sessions and the outcome variable is binary (click/no click)
for each session, which follows a binomial distribution.

@@

To our knowledge, existing approaches for sample size planning for GLMMs
have exclusively focused on planning for power. Power analysis methods
for multilevel models can be categorized into formula-based methods and
simulation-based methods
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022}). Formula-based methods rely on formulas to calculate
power directly while simulation-based methods rely on repeatedly
simulating data with a known true effect size and estimating power
empirically (i.e., what percentage of simulated datasets produces a
significant p-value). Available formula-based software packages for
multilevel models often do not include GLMMs or are limited to very
simple designs
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022};
\citeproc{ref-westfallStatisticalPowerOptimal2014}{Westfall et al.,
2014}), making it necessary to build data simulations tailored
specifically to the study design. A number of tutorials have been
published describing how to perform such simulation-based power analysis
for multilevel models
(\citeproc{ref-arendStatisticalPowerTwolevel2019}{Arend \& Schäfer,
2019}; \citeproc{ref-brysbaertPowerAnalysisEffect2018}{Brysbaert \&
Stevens, 2018};
\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}; \citeproc{ref-greenSIMRPackagePower2016a}{Green \& MacLeod,
2016}; \citeproc{ref-kainPracticalGuidePower2015}{Kain et al., 2015};
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021};
\citeproc{ref-lafitSelectionNumberParticipants2021}{Lafit et al., 2021};
\citeproc{ref-zimmerSampleSizePlanning2022}{Zimmer et al., 2022}).
However, many of these tutorials focus on linear mixed models (LMMs) and
the common study designs (but see
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021},
for a tutorial that also covers more advanced settings). This narrow
focus provides limited guidance for researchers faced with more complex
study designs, especially when little prior knowledge about plausible
effect sizes is available (see the discussion in
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}).
Simulation-based power analysis with GLMMs requires making a range of
assumptions: The (conditional) distribution assumption specifies the
distributional family for the outcome variable. Assumptions about the
random effects include the assumption of normality (i.e., that the
random effects follow a normal distribution) and the covariance
structure among the random effects (i.e., if and how they are
correlated). Making these decisions requires understanding the
underlying assumptions of the model and ensuring they align with the
characteristics of the data being analyzed. Existing tutorials often
rely on heuristics for specifying variance components (e.g., the
standard deviation of random intercepts) or assume that results from
meta-analyses or data from pilot studies are available to determine
plausible values for all model parameters. However, in practice,
knowledge about those parameters from prior studies is often limited,
which makes specifying assumptions a practical challenge (see the
discussion in (\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et
al., 2008}) and
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al.,
2021})).

@@

In Table 1, we give a short review of existing R packages that can be
used for power analysis for GLMMs.

\section{When to use tailored data
simulation?}\label{when-to-use-tailored-data-simulation}

Performing tailored simulation-based sample size planning is more
complicated and time-consuming than using the existing software tools
outlined in TABLE 1. The most important circumstances under which
tailored simulation-based sample size planning is necessary are 1)
complex study designs, 2) complex statistical hypotheses, 3) planning
for precision, 4) no available prior studies or pilot data.

>>>>

\authorsaffiliations{
{Department of Psychology, LMU Munich},{Institute of Behavioral Science
\& Technology, University of St.~Gallen},{Department of Business
Psychology, Technical University of Applied Sciences Augsburg},{UCL
Global Business School for Health, University College London}}

@@

\abstract{When planning experimental research, determining an
appropriate sample size and using suitable statistical models are
crucial for robust and informative results. The recent replication
crisis underlines the need for more rigorous statistical methodology and
adequately powered designs. Generalized linear mixed models (GLMMs)
offer a flexible statistical framework to analyze experimental data with
complex (e.g., dependent and hierarchical) data structures. However,
available methods and software for a priori sample size planning for
GLMMs are often limited to specific designs. Tailored data simulation
approaches offer a more flexible alternative. Based on a practical case
study, the current tutorial equips researchers with a step-by-step guide
and corresponding code for conducting tailored a priori sample size
planning with GLMMs. We not only focus on power analysis but also
explain how to use the precision of parameter estimates to determine
appropriate sample sizes. We conclude with an outlook on the increasing
importance of simulation-based sample size planning.}

@@

\authornote{\par{\addORCIDlink{Florian
Pargent}{0000-0002-2388-553X}}\par{\addORCIDlink{Timo K.
Koch}{0000-0001-6728-2063}}\par{\addORCIDlink{Anne-Kathrin
Kleine}{0000-0003-1919-2834}}\par{\addORCIDlink{Eva
Lermer}{0000-0002-6600-9580}}\par{\addORCIDlink{Susanne
Gaube}{0000-0002-1633-4772}}

\par{ This is version 2.5 of our preprint published at
\url{https://doi.org/10.31234/osf.io/rpjem}. All materials (reproducible
manuscript, including R code and all simulation results) are available
in the project's repository on the Open Science Framework (OSF) at
\url{https://osf.io/dhwf4/}. A Quarto Manuscripts website is hosted at
\url{https://Timo-Ko.github.io/glmm_simulation_tutorial/}.  The authors
declare that there were no conflicts of interest concerning the
authorship or the publication of this article. This research was funded
by a grant from the Volkswagen Foundation (Grant
No.~98525).   Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: Florian
Pargent:   formal
analysis, methodology, visualization, writing, editing; Timo K.
Koch:   conceptualization, formal
analysis, writing, editing; Anne-Kathrin Kleine:   formal
analysis, editing; Eva Lermer:   conceptualization, funding
acquisition, editing; Susanne
Gaube:   conceptualization, supervision, editing}
\par{Correspondence concerning this article should be addressed to Timo
K. Koch, Institute of Behavioral Science \& Technology, University of
St.~Gallen, Torstrasse
25, St.~Gallen 9000, Switzerland, Email: timo.koch@unisg.ch}
}

@@

When planning experimental research, it is essential to determine an
appropriate sample size and use appropriate statistical models to
analyze the data to ensure that the results are robust and informative
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}). The
recent replication crisis in Psychology and other disciplines has
illustrated many challenges surrounding the reproducibility and
reliability of study findings
(\citeproc{ref-yarkoniGeneralizabilityCrisis2022}{Yarkoni, 2022}). As a
result, there is a growing need for more rigorous statistical
methodology and the adoption of adequately powered experimental designs.
Multiple easy-to-use software solutions exist for simple statistical
models and experimental designs (\citeproc{ref-R-pwr}{Champely, 2020};
\citeproc{ref-lakensSimulationBasedPowerAnalysis2021}{Lakens \&
Caldwell, 2021}). A priori sample size planning for more complex
research designs such as flexible generalized linear mixed models (GLMM)
is not covered by standard software solutions. Researchers willing to
use this framework will need to use data simulation. In the present
work, we provide a tutorial on how to determine adequate sample sizes by
performing tailored simulation-based sample size planning for GLMMs.
After introducing some theoretical background on sample size planning,
we review existing software solutions in R and discuss under which
circumstances tailored data simulations are necessary. We proceed by
describing the relevant steps and decisions involved in tailored data
simulation, illustrated in a case study.

To benefit most of this tutorial paper, we recommend readers to
familiarize themselves with basic statistical concepts like hypothesis
tests (HTs) and their statistical power as well as confidence intervals
(CIs) and their precision
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021};
\citeproc{ref-lakensImprovingYourStatistical2022}{Lakens, 2022b};
\citeproc{ref-riesthuisSimulationBasedPowerAnalyses2024}{Riesthuis,
2024}). Some knowledge of causal inference is beneficial but not
necessary
(\citeproc{ref-deffnerCausalFrameworkCrossCultural2022}{Deffner et al.,
2022}; \citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al.,
2021}). In addition, readers should have an understanding of how to
conduct statistical analyses with R
(\citeproc{ref-wickhamDataScienceImport2023}{Wickham et al., 2023}) and
how to simulate data
(\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021};
\citeproc{ref-hallgrenConductingSimulationStudies2013}{Hallgren, 2013};
\citeproc{ref-leeUsingTidyversePackage2020}{Lee et al., 2020}). For data
simulation, we use functions from the \emph{tidyverse}
(\citeproc{ref-wickhamWelcomeTidyverse2019}{Wickham et al., 2019}) and
the \emph{faux} package (\citeproc{ref-R-faux}{DeBruine, 2023}).
Finally, readers should have a basic understanding of regression
modeling and GLMMs
(\citeproc{ref-brownIntroductionLinearMixedEffects2021}{Brown, 2021}).
In this tutorial, we simulate data by manually specifying the model
equation of a GLMM that represents our assumed data-generating process
(\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}). It is not necessary to understand the technical details of
how GLMMs are estimated. However, it is crucial to understand the
structure of a basic GLMM (e.g., logistic regression with random
intercepts) and how the model assumes that the dependent variable's
values are determined by the predictor variables and the random effects.

@@

Conducting research with insufficiently large sample sizes can have many
negative consequences (\citeproc{ref-buttonPowerFailureWhy2013}{Button
et al., 2013}). First, experiments may yield inconclusive or misleading
results, hindering the accumulation of knowledge. Second, studies that
are doomed never to find a postulated effect waste resources by
consuming time, effort, and funding. For these reasons, many journals
and funding bodies now require a sample size justification in study
protocols and grant proposals, recognizing its relevance in ensuring
robust and meaningful findings. While sample sizes can be justified with
resource constraints or general heuristics, statistical arguments based
on power or precision are the gold standard
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).

Most empirical studies in psychology and other social sciences apply
hypothesis testing. Consequently, the dominant approach for determining
an adequate sample size is based on power analysis (i.e., planning for
power) (\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}).
Statistical power is defined as the probability that a HT has a
significant p-value when analyzing repeated samples from a population
with a true effect of some pre-specified size
(\citeproc{ref-cohenPowerPrimer1992}{Cohen, 1992}). Less formally, power
is described as the probability that a HT correctly rejects the null
hypothesis when the alternative hypothesis is true. If the sample size
(i.e., the number of participants and/or stimuli) is insufficient to
detect the effects or relationships being investigated with high
probability, the study is considered ``underpowered''. When planning for
power, a target is set for the statistical power of a HT of interest.
Assuming an effect size of interest and a desired significance level, a
minimum sample size can be determined that, on average, would guarantee
reaching this target. The most prominent heuristic is to target a power
of \(1- \beta = 0.8\) in combination with a type I error rate of
\(\alpha = 0.05\)
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).
However, depending on the research goals or resource constraints, there
are often good reasons to move away from this standard
(\citeproc{ref-benjaminRedefineStatisticalSignificance2017}{Benjamin et
al., 2017}; \citeproc{ref-lakensJustifyYourAlpha2018}{Lakens, Adolfi, et
al., 2018}). For example, the \emph{Social Sciences Replication Project}
targeted a power of 0.9 to safeguard against biased effect sizes in the
original studies
(\citeproc{ref-camererEvaluatingReplicabilitySocial2018}{Camerer et al.,
2018}).

In contrast to power analysis, sample size planning can also be based on
the precision of parameter estimates (i.e., planning for precision or
planning for accuracy)
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}). Not
all research questions are best answered by hypothesis testing. It has
been argued that basic research rarely requires discrete decisions on
whether some effect has been ``discovered'' and should thus shift from
hypothesis testing towards an estimation framework
(\citeproc{ref-cummingNewStatisticsWhy2014}{Cumming, 2014};
\citeproc{ref-kruschkeBayesianNewStatistics2018}{Kruschke \& Liddell,
2018};
\citeproc{ref-mcelreathStatisticalRethinkingBayesian2020}{McElreath,
2020}), although this view is not without critique
(\citeproc{ref-uyguntuncEpistemicPragmaticFunction2023}{Uygun Tunç et
al., 2023}). When no HTs are conducted, power analysis is not relevant
for sample size planning. In the precision framework, the target
quantity commonly used for sample size planning is the expected width of
a CI (\citeproc{ref-kelleySampleSizePlanning2006}{Kelley \& Rausch,
2006}; \citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}). A
CI with a confidence level of 0.95 provides the smallest interval with
the property that 95\% of individual CIs would include the true quantity
of interest upon repeated sampling. Thus, a narrow CI with fewer
plausible values for the quantity of interest is more informative about
the size of the true effect than a wide CI. Apart from the confidence
level, the width of a CI depends on the sample size. Because bigger
samples carry more information, they lead to smaller CIs. When planning
for precision, a target can be set for the expected width of a CI of
interest. Assuming some effect size of interest and a certain confidence
level, a minimum sample size can be determined that would guarantee
reaching the targeted expected width. Because planning for precision is
still rare, there are no common heuristics for choosing the desired
width of the CI
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).

@@

As study designs become more complex, psychological researchers require
more sophisticated statistical models to capture their nuanced
relationships and grouping structures
(\citeproc{ref-yarkoniGeneralizabilityCrisis2022}{Yarkoni, 2022}). GLMMs
(also called multilevel models) are gaining popularity because they
offer great flexibility when applied carefully
(\citeproc{ref-brownIntroductionLinearMixedEffects2021}{Brown, 2021};
\citeproc{ref-matuschekBalancingTypeError2017}{Matuschek et al., 2017};
\citeproc{ref-meteyardBestPracticeGuidance2020a}{Meteyard \& Davies,
2020}). GLMMs are an extension of LMMs (Linear Mixed Models), which are,
in turn, extensions of linear regression models that account for
correlated data, including hierarchical structures
(\citeproc{ref-bolkerLinearGeneralizedLinear2015}{Bolker, 2015};
\citeproc{ref-fahrmeirRegressionModelsMethods2021}{Fahrmeir et al.,
2021}). In this context, correlated data means that the value in the
outcome variable for one observation may be related to the value of
another observation in a systematic way that is not already accounted
for by the usual (fixed) predictor variables (e.g., the age of
participants). This correlation can arise for various reasons: For
instance, responses to some stimuli from some participants might be more
similar because the same person was measured multiple times (repeated
measurements), participants belong to the same group (clustering), or
participants responded to the same stimulus (stimulus effects). Thus,
modeling such correlations is important whenever the data has a clear
structure, while the grouping variables can be hierarchically nested
(e.g., grouping variables students and schools: each student belongs to
exactly one school) or cross-classified (e.g., grouping variables
students and math exercises: each student is presented with several math
exercises). LMMs are used when the outcome variable is continuous and
follows a normal distribution (after conditioning on all predictor
variables). They allow for the modeling of fixed effects, which capture
the relationships between the usual predictors and the outcome, as well
as random effects, which account for the different types of correlation
structure and grouping effects. Random effects are typically assumed to
follow a normal distribution with a mean of zero and a variance that
quantifies the heterogeneity across groups. Correlated random effects
can be assumed in models that contain both random intercepts and random
slopes. GLMMs extend the LMM framework to accommodate non-normally
distributed continuous and categorical outcome variables. GLMMs involve
a link function that connects the linear combination of predictor
variables to the expected value of the outcome variable. The link
function allows for modeling the relationship between predictors and the
outcome in a non-linear way that is appropriate for the specific
distribution family of the outcome variable.

@@

To our knowledge, existing approaches for sample size planning for GLMMs
have exclusively focused on planning for power. In Table 1, we review
available software packages that can be used to perform power analysis
for multilevel models in R {[}Version 4.4.0; R Core Team
(\citeproc{ref-R-base}{2024}){]}. Power analysis methods can be
categorized into formula-based, summary-statistics-based and
simulation-based methods
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022}). Formula-based methods rely on exact formulas to
calculate power directly. Summary-statistics-based methods use
statistical theory to approximate power based on formula-based methods
developed for simple t-tests. Simulation-based methods rely on
repeatedly simulating data with a known true effect size and estimating
power empirically, that is what percentage of simulated datasets
produces a significant p-value. Available formula-based and
summary-statistics-based software packages for multilevel models often
do not include GLMMs or are limited to simple designs
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022};
\citeproc{ref-westfallStatisticalPowerOptimal2014}{Westfall et al.,
2014}), making it necessary to build data simulations tailored
specifically to the study design. A number of tutorials have been
published describing how to perform such simulation-based power analysis
for multilevel models
(\citeproc{ref-arendStatisticalPowerTwolevel2019}{Arend \& Schäfer,
2019}; \citeproc{ref-brysbaertPowerAnalysisEffect2018}{Brysbaert \&
Stevens, 2018};
\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}; \citeproc{ref-greenSIMRPackagePower2016a}{Green \& MacLeod,
2016}; \citeproc{ref-kainPracticalGuidePower2015}{Kain et al., 2015};
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021};
\citeproc{ref-lafitSelectionNumberParticipants2021}{Lafit et al., 2021};
\citeproc{ref-zimmerSampleSizePlanning2022}{Zimmer et al., 2022}).
However, many of these tutorials focus on LMMs and the most common study
designs (see \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et
al., 2021} for a tutorial that also covers more advanced settings). This
narrow focus provides limited guidance for researchers using more
complex study designs, especially when little prior knowledge about
plausible effect sizes is available (see the discussion in
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}).
Simulation-based power analysis with GLMMs requires making a range of
assumptions about the model structure that should align with the
characteristics of the data being analyzed. Existing tutorials often
rely on heuristics for specifying variance components (e.g., the
standard deviation of random intercepts) or assume that results from
meta-analyses or data from pilot studies are available to determine
plausible values for all model parameters. However, in practice,
knowledge about those parameters from prior studies is often limited,
which makes specifying assumptions a practical challenge (see the
discussion in Maxwell et al.
(\citeproc{ref-maxwellSampleSizePlanning2008}{2008}) and Kumle et al.
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{2021})). We will
discuss a number of strategies on how to specify model parameters for
application-specific, tailored data simulations in a later chapter.

@@

\section{Reasons to use tailored data
simulation}\label{reasons-to-use-tailored-data-simulation}

Performing tailored simulation-based sample size planning is more
complicated and time-consuming than using the existing software tools
outlined in TABLE 1. Circumstances under which tailored simulation-based
sample size planning is necessary include 1) complex study designs, 2)
complex statistical hypotheses, 3) planning for precision, and 4) lack
of prior studies or pilot data.

First, requirements of real-world studies are often more complex than
the simplified designs assumed by many user-friendly software packages
for sample size planning. One frequent issue in applied data analysis is
missing data (\citeproc{ref-littleStatisticalAnalysisMissing2014}{Little
\& Rubin, 2014}). Missings can be completely random (e.g., an electronic
measurement device randomly failed for technical reasons), depend on
some attributes also measured in the dataset (e.g., older subjects
refuse answering a question on income), or be caused by the measured
attribute itself (e.g., wealthy people refuse to report their income).
Moreover, many experimental designs contain conditions in which values
of the predictor variables are missing by design. This can make data
analysis more complicated because predictors have to be coded in
specific ways that prevent the estimated GLMM from becoming
unidentified. Whether missing data has an effect on the sample size
planning depends on our theoretical assumptions on how the missingness
is caused. However, it is often challenging to decide whether missing
data can be safely ignored in the data analysis and sample size planning
process based on a merely theoretical approach
(\citeproc{ref-gomilaMissingDataExperiments2022}{Gomila \& Clark,
2022}). Tailored simulation-based approaches offer the possibility to
include the assumed process of how data become missing in the data
simulation, thereby determining the required sample size based on
simulated datasets that contain missing values (for example, see
\citeproc{ref-lanePowerStrugglesEstimating2018}{Lane \& Hennes, 2018}).
As a byproduct, the simulated datasets can also be used to test whether
the intended data analysis provides the expected (unbiased) results,
despite the missing data. Although GLMMs can handle a large variety of
outcome variables, researchers are becoming increasingly aware that many
datasets might profit from even more sophisticated models. Common
examples are zero-inflated outcomes, censoring, and nonlinear predictor
effects that can be modeled with the R packages \emph{glmmTMB}
(\citeproc{ref-brooksGlmmTMBBalancesSpeed2017}{Brooks et al., 2017}) or
\emph{brms}
(\citeproc{ref-burknerAdvancedBayesianMultilevel2018}{Bürkner, 2018}).
Tailored simulation-based approaches do not share the same limitations
as the existing software solutions for power analysis that focus
exclusively on GLMMs. As long as there is a software package available
to estimate the model of interest, it is always possible to perform
tailored simulation-based sample size planning.
